---
title: "Fitabase Data Insights Report"
output: html_document
---

# Fitness Band Data Insights

This report analyzes the Fitabase data to extract patterns in physical activity, sleep, and heart rate.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(lubridate)
library(ggplot2)
library(scales)
```

## 1. Data Loading

Loading key datasets from the `data/raw/Fitabase Data 3.12.16-4.11.16` directory.

```{r load_data}
base_path <- "../data/raw/Fitabase Data 3.12.16-4.11.16/"

# Daily Activity
daily_activity <- read_csv(paste0(base_path, "dailyActivity_merged.csv")) %>%
  mutate(ActivityDate = as.Date(ActivityDate, format="%m/%d/%Y"))

# Hourly Steps
hourly_steps <- read_csv(paste0(base_path, "hourlySteps_merged.csv")) %>%
  mutate(ActivityHour = as.POSIXct(ActivityHour, format="%m/%d/%Y %I:%M:%S %p", tz="UTC"))

# Sleep Minutes
minute_sleep <- read_csv(paste0(base_path, "minuteSleep_merged.csv")) %>%
  mutate(date = as.POSIXct(date, format="%m/%d/%Y %I:%M:%S %p", tz="UTC"))

# Weight Log (Optional, check if empty)
weight_log <- read_csv(paste0(base_path, "weightLogInfo_merged.csv"))

print(paste("Daily Activity Records:", nrow(daily_activity)))
print(paste("Hourly Steps Records:", nrow(hourly_steps)))
```

## 2. Daily Activity Summary

Overview of daily physical metrics.

```{r activity_summary}
daily_summary <- daily_activity %>%
  summarise(
    Avg_Steps = mean(TotalSteps),
    Avg_Calories = mean(Calories),
    Avg_Distance = mean(TotalDistance),
    Very_Active_Mins = mean(VeryActiveMinutes),
    Sedentary_Mins = mean(SedentaryMinutes)
  )

print(daily_summary)

# Distribution of Total Steps
ggplot(daily_activity, aes(x = TotalSteps)) +
  geom_histogram(binwidth = 1000, fill = "#3498db", color = "white") +
  labs(title = "Distribution of Daily Steps", x = "Total Steps", y = "Frequency") +
  theme_minimal()
```

## 3. Hourly Activity Patterns

Identifying peak activity times during the day.

```{r hourly_patterns}
hourly_patterns <- hourly_steps %>%
  mutate(
    Hour = hour(ActivityHour),
    Weekday = wday(ActivityHour, label = TRUE, abbr = TRUE)
  ) %>%
  group_by(Weekday, Hour) %>%
  summarise(Avg_Steps = mean(StepTotal), .groups = 'drop')

# Heatmap: Day vs Hour
ggplot(hourly_patterns, aes(x = Hour, y = Weekday, fill = Avg_Steps)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "#e74c3c") +
  labs(title = "Average Hourly Steps Heatmap", x = "Hour of Day", y = "Day of Week") +
  theme_minimal() +
  scale_x_continuous(breaks = seq(0, 23, 2))
```

## 4. Sleep Analysis

Aggregating minute-level sleep data to daily duration.

```{r sleep_analysis}
daily_sleep <- minute_sleep %>%
  mutate(Date = as.Date(date)) %>%
  group_by(Id, Date) %>%
  summarise(
    Sleep_Minutes = n(),
    Sleep_Hours = n() / 60,
    .groups = 'drop'
  )

# Sleep Duration Histogram
ggplot(daily_sleep, aes(x = Sleep_Hours)) +
  geom_histogram(binwidth = 0.5, fill = "#8e44ad", color = "white") +
  labs(title = "Distribution of Sleep Duration", x = "Sleep (Hours)", y = "Count") +
  geom_vline(xintercept = 7, linetype="dashed", color="red", size=(1)) +
  annotate("text", x=7.2, y=5, label="Rec. 7h", color="red", angle=90) +
  theme_minimal()
```

## 5. Correlations

Is there a relation between how active you are and how well you sleep?

```{r correlations}
merged_data <- inner_join(daily_activity, daily_sleep, by = c("Id", "ActivityDate" = "Date"))

# Steps vs Sleep
ggplot(merged_data, aes(x = TotalSteps, y = Sleep_Hours)) +
  geom_point(alpha = 0.6, color = "#2c3e50") +
  geom_smooth(method = "lm", color = "#e67e22") +
  labs(title = "Correlation: Daily Steps vs Sleep Duration", x = "Daily Steps", y = "Sleep Hours") +
  theme_minimal()

cor_val <- cor(merged_data$TotalSteps, merged_data$Sleep_Hours)
print(paste("Correlation between Steps and Sleep:", round(cor_val, 3)))
```

## 6. User vs. Ideal Comparison

Benchmarking the average user in this dataset against health standards.

```{r ideal_comparison}
# Calculate User Profile
user_profile <- merged_data %>%
  summarise(
    Steps = mean(TotalSteps),
    Calories = mean(Calories),
    Sleep_Hours = mean(Sleep_Hours)
  ) %>%
  mutate(Type = "User Average")

# Define Ideal Profile
ideal_profile <- data.frame(
  Steps = 10000,
  Calories = 2500,
  Sleep_Hours = 8,
  Type = "Ideal Healthy"
)

comparison <- bind_rows(user_profile, ideal_profile) %>%
  pivot_longer(cols = c(Steps, Calories, Sleep_Hours), names_to = "Metric", values_to = "Value")

# Plot
ggplot(comparison, aes(x = Type, y = Value, fill = Type)) +
  geom_bar(stat = "identity", position = "dodge", width = 0.7) +
  facet_wrap(~Metric, scales = "free_y") +
  theme_minimal() +
  labs(title = "User vs. Ideal Benchmark", y = "Value", x = "") +
  scale_fill_manual(values = c("User Average" = "#3498db", "Ideal Healthy" = "#2ecc71"))
```
